---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```
# Tools for modeling bumblebee colony growth

<!-- badges: start -->
[![R build status](https://github.com/Aariq/bumbl/workflows/R-CMD-check/badge.svg)](https://github.com/Aariq/bumbl/actions)
[![Codecov test coverage](https://codecov.io/gh/Aariq/bumbl/branch/master/graph/badge.svg)](https://codecov.io/gh/Aariq/bumbl?branch=master)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
[![CRAN status](https://www.r-pkg.org/badges/version/bumbl)](https://CRAN.R-project.org/package=bumbl)

<!-- badges: end -->

`bumbl` implements a model for bumblebee colony growth described in Crone and Williams 2016^1^. The `brkpt` function models colony growth as having a change point at some time, *tau*, where the colony switches from growth and worker production to gyne production.  The `bumbl` function applies this model to a data frame of data from multiple colonies, allowing for each colony to have it's own *tau* and returns the original data augmented with coefficients from the changepoint model.

This is still in very early development, so use at your own risk.

More bumblebee related functions to come...

## Installation

You can install this development version of `bumbl` with:

``` r
devtools::install_github("Aariq/bumbl", build_vignettes = TRUE)
```

## Getting started

View the package vignette with:

``` r
library(bumbl)
vignette("bumbl")
```
```{r include=FALSE}
# library(bumbl)
devtools::load_all()
```

View the `bomubs` dataset

```{r}
head(bombus)
```

### Example use

Using a subset of the `bombus` dataframe to estimate the week (tau) that colonies switch to reproduction

```{r}
bombus2 <- bombus[bombus$colony %in% c(9, 82, 98, 35), ]
results <- bumbl(bombus2, colonyID = colony, t = week, formula = d.mass ~ week)
results
```

Plot the results

```{r}
par(mfrow = c(2, 2))
plot(results)
par(mfrow = c(1, 1))
```


# References
^1^Crone, E. E., and Williams, N. M. (2016). Bumble bee colony dynamics: quantifying the importance of land use and floral resources for colony growth and queen production. Ecol. Lett. 19, 460–468. https://doi.org/10.1111/ele.12581

