---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```
# Tools for modeling bumblebee colony growth

<!-- badges: start -->
[![R build status](https://github.com/Aariq/bumbl/workflows/R-CMD-check/badge.svg)](https://github.com/Aariq/bumbl/actions)
[![Codecov test coverage](https://codecov.io/gh/Aariq/bumbl/branch/master/graph/badge.svg)](https://codecov.io/gh/Aariq/bumbl?branch=master)
[![Lifecycle: experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
[![CRAN status](https://www.r-pkg.org/badges/version/bumbl)](https://CRAN.R-project.org/package=bumbl)
[![DOI](https://zenodo.org/badge/200284602.svg)](https://zenodo.org/badge/latestdoi/200284602)
<!-- badges: end -->

`bumbl` implements a model for bumblebee colony growth described in Crone and Williams 2016^1^. It models colony growth as having a change point at some time, *tau*, where the colony switches from growth and worker production to gyne production and decline.  Currently `bumbl()` works by fitting a separate changepoint model to each colony, optimizing the switchpoint, *tau*. It returns the optimal switchpoint, growth, and decline rates for each colony. Becuase the current version of `bumbl()` works by fitting a separate GLM to each colony, if covariates are included, their estimates could vary significantly among colonies.  Stay tuned for future developments that may allow estimating a single value of a covariate but different values of growth and decline rates and *tau* for each colony.

This is still in early development, so use at your own risk. 

## Collaboration

I'm looking for collaborators who know (or are willing to let me teach them) how to use git and GitHub and who have an interest in helping to develop and maintain this package long-term.  I'm not a bumblebee biologist, so I would especially love a collaborator who works on bumblebees or other organisms with a similar growth, switch, decline lifecycle.

I also welcome contributions including bug-fixes, improvement of documentation, additional features, or new functions relating to bumblebee ecology and demography from anyone!

### Roadmap

  - [x] Write `bumbl()` function to model colony growth with switchpoint
  - [x] Finish documentation and vignette(s)
  - [x] [rOpenSci
    review](https://github.com/ropenscilabs/statistical-software-review/issues/2)
  - [x] Regroup with Elizabeth Crone to discuss current behavior of `bumbl()`
  - [x] Release v0.1.0 and archive on Zenodo
  - [ ] Possibly re-work internals of `bumbl()`? (see [#58](https://github.com/Aariq/bumbl/issues/58))
  - [ ] Polish user interface in regards to specifying formula and covariates.
  - [ ] Submit package to CRAN
  - [ ] Write a manuscript
  
Possible areas of improvement that I don't personally have time for:

- Extend `bumbl()` to work with GLMMs


## Installation

You can install this development version of `bumbl` with:

``` r
devtools::install_github("Aariq/bumbl", build_vignettes = TRUE)
```

## Getting started

View the package vignette with:

``` r
library(bumbl)
vignette("bumbl")
```
```{r include=FALSE}
# library(bumbl)
devtools::load_all()
```

View the `bomubs` dataset

```{r}
head(bombus)
```

### Example use

Using a subset of the `bombus` dataframe to estimate the week (tau) that colonies switch to reproduction

```{r}
bombus2 <- bombus[bombus$colony %in% c(9, 82, 98, 35), ]
results <- bumbl(bombus2, colonyID = colony, t = week, formula = d.mass ~ week)
results
```

Plot the results

```{r}
par(mfrow = c(2, 2))
plot(results)
par(mfrow = c(1, 1))
```


# References
^1^Crone, E. E., and Williams, N. M. (2016). Bumble bee colony dynamics: quantifying the importance of land use and floral resources for colony growth and queen production. Ecol. Lett. 19, 460–468. https://doi.org/10.1111/ele.12581


---


Please note that the bumbl project is released with a
[Contributor Code of Conduct](CODE_OF_CONDUCT.md). By contributing to this
project you agree to abide by its terms.

