---
title: "Modeling bumblebee colony growth with a switch point"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Modeling bumblebee colony growth with a switch point}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(bumbl)
library(car)
```

## Introduction to the bumbl() function

The `bumbl()` function fits a model that allows for different growth rates before and after a switch point, $\tau$, and an added decay rate after $\tau$. This model is fit separately for each colony in a dataset.  In this example, we'll use the built-in `bombus` dataset.  For more information on this dataset see the help file with `?bombus`.

Before the switch point, growth (colony weight, $W_t$) is defined as:

$$
W_t = \lambda^tW_0
$$
After the switchpoint ($t > \tau$), it's defined as:

$$
W_t = \lambda^{\tau}W_{0}\delta^{t-\tau}
$$

Where $\delta$ is a rate of decline.  Therefore:

$$
W_t = 
  \begin{cases}
  \lambda^tW_0 & t\leq\tau \\
  \lambda^{\tau}W_{0}\delta^{t-\tau} & t > \tau
  \end{cases}
$$


After log-linearizing it it looks like this:

For $t<=\tau$:

$$
\ln(W_t) = \ln(W_0) + t\ln(\lambda)
$$


For $t>\tau$

$$
\ln(W_t) = \ln(W_0) + \tau\ln(\lambda) + (t-\tau)\ln(\delta)
$$

Turning that into a simple linter regression looks like:

$$
ln(W_t) = \beta_0 + \beta_1*t + \beta_2*post
$$

Where *post* is $0$ before $\tau$ and $t-\tau$ after $\tau$

To clarify, in the above equation:

- $\beta_0 = \ln(W_0)$
- $\beta_1 = \ln(\lambda)$
- $\beta_2 = \ln(\delta-\lambda) = \frac{\ln(\delta)}{\ln(\lambda)}$

## Modeling Bombus colony growth

```{r}
head(bombus)
```

The `bumbl()` function expects a tidy dataset with a single column for colony ID, a single column for some measure of time, and a single column for some measure of colony growth at minimum.

```{r}
out <- bumbl(bombus, colonyID = colony, t = week, formula = log(mass) ~ week)
head(out)
```

For each colony, we get the maximum likelihood estimate for $\tau$ (`tau`), the estimated initial colony weight ($\ln({W_0})$, `logN0`) on a log scale, the average colony growth rate ($\ln(\lambda)$, `logLam`) on a log scale, the rate of decline after $\tau$ ($ln(\delta - \lambda)$, `decay`), and the estimated maximum weight of each colony, on a log scale (`logNmax`).

You can also include co-variates 
TODO:[why might you want to do this?]

```{r}
out2 <- bumbl(bombus, colonyID = colony, t = week, formula = log(mass) ~ week * cum_floral)
head(out2)
```


## Plotting results

To check that these results are sensible, you should plot them.  To do this, you first need an "augmented" dataframe that you can get using the `augment = TRUE` argument in `bumbl()`.

```{r}
#using a subset of bomubs for this
bombus2 <- bombus[bombus$colony %in% c(38, 82, 98, 35), ]

aug_out <- bumbl(bombus2, colonyID = colony, t = week,
                 formula = log(mass) ~ week, 
                 augment = TRUE)
head(aug_out)
```

Then you can plot this dataframe with `bumbl_plot()`.

```{r}
bumbl_plot(aug_out)
```

From this it's clear that for colony 38, the model is incorrect.


## Using results

With `augment = TRUE`, `bumbl()` returns your original data with columns added for model coefficients.  This might be useful if you were interested in how habitat type effects the switch-time, $tau$.

```{r}
out3 <- bumbl(bombus, colonyID = colony, t = week, formula = log(mass) ~ week, augment = TRUE)
head(out3)
```

We can now use that augmented dataset in another model

```{r}
m <- lm(tau ~ habitat, data = out3)
summary(m)
Anova(m)
```

In this example, without accounting for any other covariates, habitat type has a highly significant effect on switch time with colonies in organic farms switching to decline about two weeks earlier than conventional agriculture and about a week earlier than wild landscapes. 



