---
title: "Colony Growth Model"
author: "Eric R. Scott"
date: "2019-07-31"
output: 
  html_document: 
    highlight: kate
    theme: yeti
    toc: yes
    toc_float: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

*Last compiled: `r Sys.Date()`*

```{r packages, include=FALSE}
library(tidyverse)
library(here)
library(bumbl)
```

# Purpose

Try to get Elizabeth Crone's breakpoint bumble bee colony growth model working on my machine

# Load Data

```{r data, echo=TRUE}
bombus
```


## Data Dictionary




# Notes

The idea is to fit a model that allows for different growth rates before and after a switch point, $\tau$, and an added decay rate after $\tau$.

Before the switch point, growth (colony weight, $W_t$) is defined as:

$$
W_t = \lambda^tW_0
$$
After the switchpoint ($t > \tau$), it's defined as:

$$
W_t = \lambda^{\tau}W_{0}\delta^{t-\tau}
$$
Where $\delta$ is a rate of decline

$$
W_t = 
  \begin{cases}
  \lambda^tW_0 & t\leq\tau \\
  \lambda^{\tau}W_{0}\delta^{t-\tau} & t > \tau
  \end{cases}
$$


After log-linearizing it it looks like this:

For $t<=\tau$:

$$
\ln(W_t) = \ln(W_0) + t\ln(\lambda)
$$

For $t>\tau$

$$
\ln(W_t) = \ln(W_0) + \tau\ln(\lambda) + (t-\tau)\ln(\delta)
$$
Turning that into a simple linter regression looks like:
$$
ln(W_t) = \beta_0 + \beta_1*t + \beta_2*post
$$
Where *post* is $0$ before $\tau$ and $t-\tau$ after $\tau$

To clarify:

- $\beta_0 = \ln(W_0)$
- $\beta_1 = \ln(\lambda)$
- $\beta_2 = \ln(\delta-\lambda) = \frac{\ln(\delta)}{\ln(\lambda)}$

# Original code dissection

Elizabeth's code contained an additional co-variate, "supp" which was whether hives were supplemented or not.  But for the sake of writing a working function, I'll ignore covariates for now.

What the function needs to do is for each colony:

1. calculate *post*
2. Fit a linear model, `lm(log(TrueColonyWt_g) ~ Round + post, data = dat)`, using a range of $\tau$s
3. Find the best $\tau$ using maximum liklihood
4. Export coefficients from the model


```{r}
taos = (seq(2,8,0.1)) #possible values for breakpoint
taoML = array() 
sites = unique(wts$ColonyID) 
for(j in 1:length(sites)){
  dat = wts[wts$ColonyID == sites[j],] #subset data by colony?
  LLs = array(NA, dim = c(length(taos),1)) #creates array with slot for each tau?
  
  for(i in 1:length(taos)){
    usetao = taos[i]
    dat$post = dat$Round - usetao
    dat$post[(dat$post < 0) == T] = 0 #post is 0 if before tao, numeric if after tao
    
    m0 = try(lm(log(TrueColonyWt_g) ~ Round + post, data = dat))
    if(class(m0) != "try-error") LLs[i] = logLik(m0)
    LLs
  }
  taoML[j] = taos[which(LLs == max(LLs))]
  
}
taoML # show the list of maximum-likelihood breakpoints for each colony
```
I think the warning is because sometimes there are multiple quivalent taus with the same maximum liklihood.  That is, `taos[which(LLs == max(LLs))]` is sometimes a vector.


Next, loop below extracts the colony-dynamic coefficients for each colony, given the ML value of tao.  It **re-fits** the model using the best $\tau$, which is probably not the best way.

```{r}

params = array(NA, dim = c(length(sites), 6)) #<-had to change this because I got rid of "supp"
for(k in 1:length(sites)){
  dat = wts[wts$ColonyID == sites[k],]
  usetao = taoML[k] #isn't this going to be a vector sometimes?
  dat$post = dat$Round - usetao
  dat$post[(dat$post < 0) == T] = 0
  m1 = lm(log(TrueColonyWt_g) ~ Round + post, data = dat) 
  print(sites[k])
  print(summary(m1))
  params[k,] = (c(sites[k],usetao, coef(m1), max(predict(m1), na.rm = T)))
  # plots of each colony's trajectory and fitted curve, for visual inspection
  # these lines of code can be turned off without loss of function
  # but I like to look at them to be sure nothing went wrong
  plot(dat$Round, dat$TrueColonyWt_g, main = sites[k])
  points(dat$Round[1:length(predict(m1))], exp(predict(m1)), type = "l", col = "red")
}# this loop returns `params` and prints models and plots as side-effect


```
First problem: This doesn't allow for variable numbers of parameters.  It fails because it was expecting 7 parameters I think.


Next, this code converts the matrix of model parameters to a data frame
```{r}
params = data.frame(params)
params
colnames(params) = c("ColonyID", "tao", "logNo", "loglam", "decay", "logNmax") #can't just assign these manually like this.
toc()
head(params)
```

- `tao` is the switchpoint, in weeks.  The colony grows for tao weeks, then begins to decline in week tao + 1
- `logNo` is the intercept of the growth function.  It reflects actual initial colony size, if the colony initially grows exponentiallyl.  It would also be lower if there were a few weeks lag before growth started in the field
- `loglam` is the average 10-day (log-scale) colony growth rate (i.e., rate of weight gain) during the growth period
- other coefs for covariates
- `decay` reflects the rate of decline during the decline period.  In fact, the way this model is set up, the actual  rate of decline per 10 days is calculated as decay - loglam.
- `logNmax` is the maximum weight reached by each colony.  It is a function of tao, logNo and loglam

And this appends it back to the original data, but summarized by colony and all the co-variates in the model
```{r}
param.dat = merge(params, wts[wts$Round == 0,c(2,4,5)], all.y = F) #full_join()
param.dat = param.dat[1:length(sites), ] #not sure why necessary
param.dat
#maximum-likelihood breakpoints for each colony
```


# Functionalizing

## Find Tau for a single colony
First, let's write a function that picks the best tau for *one* colony

```{r}
library(bumbl)
testbees <- bombus %>% filter(colony == 68)
brkpt(testbees, t = week, formula = log(mass) ~ week)
```

## Map brkpt to all colonies

Using possibly() with nested data works great at getting error mesages with readable compact code.
BUT I'll have to use something more complicated like quietly

```{r}
# safe_brkpt <- safely(brkpt)
safe_brkpt <- possibly(brkpt, NULL)

test <- 
  bombus %>% 
  group_by(colony) %>% 
  nest() %>% 
  mutate(safe_model = map(data,
                     ~safe_brkpt(.,
                            t = week,
                            # taus = seq(1:30, 0.5),
                            formula = log(mass) ~ week))) %>% 
  unnest(safe_model, keep_empty = TRUE) 

```

```{r}
errs <- test %>% filter(is.null(model[[1]]))
if(nrow(errs) > 0){
  abort(paste0("More than 1 equivalent tau for colony ID ", str_flatten(errs$colony, ", ")))
}
colonyID <- quo(colony)
t <- quo(week)
test %>% 
  mutate(coefs = map(.data$model, broom::tidy)) %>%
    unnest(.data$coefs) %>%
    select(!!colonyID, "tau", "model", "term", "estimate") %>%
    spread(key = "term", value = "estimate") %>%
    mutate(logNmax = map_dbl(.data$model, ~max(predict(.), na.rm = TRUE))) %>%
    select(-"model") %>%
    select(!!colonyID, "tau", logNo = '(Intercept)', loglam = !!t, decay = '.post', everything())
```
```{r}
library(forcats)
library(purrr)

mydata <- colony_weights %>% 
  dplyr::filter(!ColonyID %in% c("68", "97")) 

dflist <-
  mydata %>% 
  group_by(ColonyID) %>%
  group_split(keep = TRUE) %>%  #experimental
  set_names(pull(group_keys(mydata, ColonyID)))
```
```{r}
#for loop style
model_list <- vector("list", length(dflist))


for(i in 1:length(dflist)){

  model_list[[i]] <- 
    tryCatch(brkpt(dflist[[i]], taus = mytaus,t = Round,formula = log(TrueColonyWt_g) ~ Round),
             error = function(c){
               c$message <- glue::glue("For Colony ID '{names(dflist)[i]}': {c$message}")
               stop(c)
             })
  names(model_list) <- names(dflist)
}
bind_rows(model_list, .id = "Colony ID")

```
 
```{r}
dflist %>% #experimental
  map_dfr(~brkpt(.x,
                taus = mytaus,
                t = Round,
                formula = log(TrueColonyWt_g) ~ Round), .id = "ColonyID") %>% 
  mutate(coefs = map(model, broom::tidy)) %>%
  unnest(coefs, .preserve = "model") %>% 
  select(ColonyID, tau, model, term, estimate) %>% 
  spread(key = term, value = estimate) %>%
  mutate(logNmax = map_dbl(model, ~max(predict(.), na.rm = TRUE))) %>% 
  select(-model) %>%
  select(ColonyID, tau, logNo = `(Intercept)`, loglam = Round, decay = .post, everything())
full_join(mydata, summarydata)
```

## How to catch errors correctly
What I want is for errors generated by `brkpt()` to be passed on to bumbl, but with info about which colonyID caused the error.

One possibility is to switch `bumbl` to a for-loop, but that seems silly just to get the errors right

Another possibility is to capture errors with `safely()` or `try()` or something in the dataframe and then construct meaningful errors inside `bumbl`

Another possibility is I'm just doing something wrong in how I pass variables from `bumbl` to `brkpt`

Code for aggregating warnings:
```{r}
at_least_one_warning <- function(code) {
  warnings <- character()
  wHandler <- function(w) {
    warnings <<- c(warnings, w$message)
    invokeRestart("muffleWarning")
  }
  
  result <- withCallingHandlers(
    expr = code,
    warning = wHandler
  )
  
  if (length(warnings) > 0) {
    message("at least one warning")
  }
  
  result
}

at_least_one_warning(purrr::map_dbl(c(10, 12, -1, -12), log))
```

from https://community.rstudio.com/t/quickly-scanning-purrr-side-effects-and-using-pillar-to-style-the-report/17075
```{r}
quietly_status = function(x, .prefix_err = 'Error:', .prefix_warn = 'Warning:',
  .prefix_msg = 'Message:', .text_ok = 'OK') {
  map(x,
      ~ if (is.null(.$warning) | is_empty(.$warning)) {
        if (is.null(.$message) | is_empty(.$message)) {
          .text_ok
        } else { paste(.prefix_msg, .$message) }
      } else { paste(.prefix_warn, .$warning) }) %>%
  as_vector()
}


# safe_log = safely(log)
quiet_log = quietly(log)

test = 
  # tidy up and trim down for the example
  mtcars %>%
  rownames_to_column(var = "car") %>%
  as_data_frame() %>%
  select(car, cyl, disp, wt) %>%
  # spike some rows in cyl == 4 to make them fail
  mutate(wt = case_when(
    wt < 2 ~ -wt,
    TRUE ~ wt)) %>%
  # nest and do some operations quietly()
  nest(-cyl) %>%
  mutate(
    qlog = map(data, ~ quiet_log(.$wt)),
    qlog_status = quietly_status(qlog),
    # optional: overwrite the "quiet" output with just successful results
    qlog = map(qlog, "result"))

test
```

THIS WORKS AND DOES WHAT I WANT.  Captures errors **AND** warnings, in a tidy format, and allows filtering and passing on of error messages.  I don't know if it's the best way, but I should do it like this for now.

```{r}

quietly_status = function(x, .prefix_err = 'Error:', .prefix_warn = 'Warning:',
  .prefix_msg = 'Message:', .text_ok = 'OK') {
  map(x,
      ~ if (is.null(.$error) | is_empty(.$error))   {
        if (is.null(.$warning) | is_empty(.$warning)) {
          if (is.null(.$message) | is_empty(.$message)) {
            
            .text_ok
            
          } else { paste(.prefix_msg, .$message) }
        } else { paste(.prefix_warn, .$warning) }
      } else { paste(.prefix_err, .$error) }) %>%
    as_vector()
}


safely_n_quietly <- function(.f, otherwise = NULL) {
  retfun <- quietly(safely(.f, otherwise = otherwise, quiet = FALSE))
  function(...) {
    ret <- retfun(...)
    list(result = ret$result$result,
         output = ret$output,
         messages = ret$messages,
         warnings = ret$warnings,
         error = ret$result$error)
  }
}

weirdlog <- function(x) {
  out <- vector(length = length(x))
  for(i in 1:length(x)) {
    if(x[i] == 0) {
      stop("I hate zero")
    } else {
      out[i] <- log(x[i])
    }
  }
  return(out)
}
weirdlog(c(1,2,3, 0))
safe_n_quiet_log <- safely_n_quietly(weirdlog)

safe_n_quiet_log(0)
quiet_log <- quietly(weirdlog)
safely_log <- safely(weirdlog)

test = 
  # tidy up and trim down for the example
  mtcars %>%
  rownames_to_column(var = "car") %>%
  as_data_frame() %>%
  select(car, cyl, disp, wt) %>%
  # spike some rows in cyl == 4 to make them fail
  mutate(wt = case_when(wt < 1.3 ~ -wt,
                        wt > 3.5 ~ 0,
                        TRUE ~ wt)) %>%
  # nest and do some operations quietly()
  nest(-cyl) %>%
  mutate(qlog = map(data, ~ safe_n_quiet_log(.$wt)),
         qlog_status = quietly_status(qlog),
         # optional: overwrite the "quiet" output with just successful results
         qlog = map(qlog, "result"))

test

log(c(1,2,3))

```

