---
title: "Single-model rework of bumbl(), to handle covariates better"
author: "Eric R. Scott"
date: "2021-02-17"
output: 
  html_document: 
    highlight: kate
    theme: yeti
    toc: yes
    toc_float: yes
    number_sections: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(here)
library(conflicted)
library(bumbl)

conflict_prefer("lag", "dplyr")
conflict_prefer("filter", "dplyr")
```

```{css}
.alert {
  position: relative;
  padding: .75rem 1.25rem;
  margin-bottom: 1rem;
  background: #f6fbfe;
  color: #12537e;
  border: 1px solid transparent;
  border-color: #1976d2;
  border-radius: .25rem;
  font-size: inherit;
}
.center {
  text-align: center;
}
```

*Last compiled: `r Sys.Date()`*

# Purpose

Analysis code demonstrating the proposed changes to how `bumbl()` works.

# Load Data

```{r data, echo=TRUE}
bombus_sub <-
  bombus %>% 
  filter(colony %in% c(9, 17, 20))
```

# Status quo

The way bumbl currently works is on one colony at a time.
The data and formula for a GLM are modified like so:

```{r}
#formula supplied by user
f_user <- d.mass ~ week

#a guess for tau, will discuss optimizing later
tau <- 5

#just a single colony's data
bombus_9 <- bombus_sub %>% filter(colony == 9)

#mutate data
df_new <-
  bombus_9 %>% 
  mutate(.post = ifelse(week <= tau, 0, week - tau))

#alter formula
f_internal <- update(f_user, .~. + .post)
f_internal
```

Then we fit the GLM

```{r}
glm(formula = f_internal, data = df_new, family = gaussian(link = "log"))
```

To find the correct value for `tau`, we need to find the value that maximizes likelihood for this model.
This is all done by `brkpt()`, a lower-level function that `bumbl()` calls for each colony in a dataset.

```{r}
bumbl:::brkpt(bombus_9, t = week, formula = d.mass ~ week, family = gaussian(link = "log"))
```

Then `bumbl` extracts some things from the model for each colony

```{r}
bumbl(
  bombus_sub,
  colonyID = colony,
  t = week,
  formula = d.mass ~ week,
  family = gaussian(link = "log")
)
```

## Covariates

In the current version of `bumbl`, covariates *can* be added to the model formula, but they are estimated for each colony separately.

```{r}
bumbl(
  bombus_sub,
  colonyID = colony,
  t = week,
  formula = d.mass ~ week + cum_floral
)
```

If you wanted to estimate a single coefficient for a covariate, the current workaround is to first use `bumbl` to find the optimal value for `tau` without the covariate, then re-fit the model "manually".

```{r}
out_df <-
  bumbl(
  bombus_sub,
  colonyID = colony,
  t = week,
  formula = d.mass ~ week,
  family = gaussian(link = "log"),
  augment = TRUE
)
out_df <-
  out_df %>% 
  mutate(.post = ifelse(week <= tau, 0, week - tau))

glm(d.mass ~ week + .post + cum_floral, data = out_df, family = gaussian(link = "log"))
```

This workaround could easily be formalized into `bumbl()` by allowing the user to override the optimization step and supply their own values of `tau`.

# New Behavior

If covariates should be taken into account **during optimization of tau**, but only one coefficient is to be estimated (not one per colony), then this requires a change in the way `bumbl()` works.

I think the new base model it needs to fit is:

    d.mass ~ week*colony + .post*colony

So that a different growth and decay rate can be fit for each colony.
`tau` then needs to be optimized for all colonies simultaneously, since `tau` determines the value of `.post`.
Then the user has more flexibility in specifying covariates.
Take a simple categorical covariate like `habitat`.
One can add `+ habitat` to estimate different intercepts for different habitats, `+ habitat*week` to have different growth rates in different habitats, or `+ habitat*.post` to have different decay rates in different habitats (how the API will work is up for discussion, but maybe `bumbl()` will accept a right-sided formula for covariates).
Importantly, these covariates are estimated across all colonies, not for each colony independently.

## Analysis code

Here's an analysis with 3 colonies and a covariate (no interactions with the covariate).

```{r}
taus <- c(5, 6, 7) #example guesses for tau
colonies <- unique(bombus_sub$colony)
bombus_sub <- 
  bombus_sub %>% 
  mutate(tau = case_when(
    colony == colonies[1] ~ taus[1],
    colony == colonies[2] ~ taus[2],
    colony == colonies[3] ~ taus[3]
  )) %>% 
  mutate(.post = ifelse(week <= tau, 0, week - tau))
m <-
  glm(
    d.mass ~ week * colony + .post * colony + habitat,
    family = gaussian(link = "log"),
    data = bombus_sub
  )
coef(m)
```

```{r results="asis"}
library(equatiomatic)
extract_eq(m, wrap = TRUE)
```

There is a different intercept (e.g. $\beta_2$), growth rate slope (e.g. $\beta_6$), and decline rate slope (e.g. $\beta_8$) for each colony as well as an overall different intercept for the two `habitat` types ($\beta_5$).

## Optimizing multiple taus

This is still a long way off from being implemented into `bumbl()`, but it's a proof of concept.

First, create a function.

```{r}
foo <- function(taus) {
  colonies <- unique(bombus_sub$colony)
  bombus_sub <- 
    bombus_sub %>% 
    # This mutate just adds a column of taus giving each colony a different tau.
    # There is probably a more efficient way of doing this for any number of
    # colonies, but doing this programmatically is key to this whole thing
    # working.
    mutate(tau = case_when(
      colony == colonies[1] ~ taus[1],
      colony == colonies[2] ~ taus[2],
      colony == colonies[3] ~ taus[3]
    )) %>% 
    # Create .post column, as before
    mutate(.post = ifelse(week <= tau, 0, week - tau))
  # Fit the model
  m <-
    glm(
      d.mass ~ week * colony + .post * colony + habitat,
      family = gaussian(link = "log"),
      data = bombus_sub
    )
  return(m)
}

foo(taus = c(5, 6, 7))
```

The `foo()` function only works with three colonies, so you'd need to edit it to work with a variable number of colonies.
Now, for `optim()` to work on the `foo()` function, it needs to be wrapped such that the parameters (`taus`) are the first argument and it returns the model log-likelihood (or whatever you want to minimize/maximize).

```{r}
foo_wrap <- function(taus) {
  m <- foo(taus = taus)
  return(logLik(m))
}
foo_wrap(taus = c(5, 6, 7))
```

Ok, now I can try to `optim()`

::: {.alert}
::: {.center}
**NOTE**
:::

There are different methods for optimization that can be supplied with the `method` argument to `optim()`.
This is just the default for demonstration, but it is probably worth while reading more about the options.
Some, like `"L-BFGS-B"` take a lower and upper bound, which might be appropriate for `tau`.
:::

```{r}
out <- 
  optim(
    par = c(5,5,5), #must supply initial values.  Could be mean(week)?
    fn = foo_wrap,
    control = list(fnscale = -1) #include to maximize instead of minimize logLik
  )
out$par
```

These are the optimal values of tau for the three colonies.

Now we can run the model with the optimal taus by feeding them back into `foo()`

```{r}
foo(out$par)
```

The values are very similar to what we get from `bumbl()`, but now habitat is included as a covariate.

For example, the `logN0` for colony 17 is 3.388 (the intercept), for colony 20 its 3.388-0.599 = 2.739.
The corresponding values from `bumbl()` without habitat as a covariate are 3.35 and 2.8.

```{r}
bumbl(
  bombus_sub,
  colonyID = colony,
  t = week,
  formula = d.mass ~ week,
  family = gaussian(link = "log")
)
```

If I'm thinking about this correctly, it seems promising.
The big hurdles are in the programming and designing the user interface.
How should the user supply the glm formula?
Something like this?

``` {.r}
bumbl(
  response = d.mass,
  time = week,
  colonyID = colony,
  covariates  = ~ habitat + cum_floral*week
)
```

But then what does `cum_floral*week` do?
Does it *actually* do `~cum_floral*week + cum_floral*.post`?
